What We Eval in the Shadows

### 2021-04-17 paper/main.tex
Updates in 6

  \item \emph{Assignments \c{(<-)}} appear in the cases where a developer wants
    to create or update a binding in a specific environment where either the
    name or the value come from some expression. Most cases originate from trivial
    code generation where the assign expression is assembled using \c{parse} or
    \c{substitute}, often in a loop: \lstinline|eval(parse(text=paste(capture.output(default_args[[i]]), "<-", i)))|.

@@

  \item \emph{Function calls \c{(F(), F(X), F(F(X)))}} represent two very
    frequent patterns: \emph{code generation} and \emph{call reflection}. There are numerous
    cases for code generation. One occurring usage is to generate ad-hoc calls
    to libraries such as \c{ggplot2} when one needs to parameterize data or
    aesthetics selection from a variable. For example \lstinline|ggplot(df, aes(x=A, y=B))|
    maps X and Y axes in the plot to \c{A} and \c{B} columns in
    \c{df}. It might not be immediately clear\footnote{\cf
    \url{https://stackoverflow.com/questions/22309285/how-to-use-a-variable-to-specify-column-name-in-ggplot}}  for a ggplot2 user
    how to modify the call to pick the actual mapping from \c{A} and \c{B}
    variables. Another case is the traditional use of \eval as a means to
    reduce boilerplate code; simple and repetitive code can easily be replaced
    with judicious use of \eval. For example, the \c{data.table} package uses
    \eval to call the \c{options} function with named arguments taken from a
    vector of strings. While the benefits are limited in this example, it is an
    attractive tool for programmers.
    %
    \begin{lstlisting}
    opts = c("datatable.verbose"="FALSE", # ...many others
    for (i in names(opts)) eval(parse(text=paste0("options(",i,"=",opts[i],")")))
    \end{lstlisting}

    Call reflection refers to the combination of \eval with \c{match.call}, a
    function that returns the current call with all its arguments captured as
    expressions. This is used mostly in the following circumstances: (1) to
    record a call for a later use in a possibly different environment, (2) pass
    most of the captured call arguments to another function or (3) evaluate
    arguments of the function in different environments. This is widely used in
    relation with statistical modeling in combination with a model fitting
    function or with \c{model.frame} and related functions. For example the
    \c{survival::coxph} function\footnote{Function fitting a Cox proportional hazards
    regression model.} contains:
    %
    \begin{lstlisting}
    Call <- match.call()
    tform <- Call[c(1,indx)]  # only keep the arguments we wanted
    tform[[1L]] <- quote(stats::model.frame)  # change the function called
    mf <- eval(tform, parent.frame())
    \end{lstlisting}

    \todo{Add that this pattern is advised by the R documentation of match.call}

@@

\subsection{Domain Specific Language.} R expressions are built and transformed symbolically and then evalled to a concrete value. For instance,   base R packages \emph{stats} provides two operators for \emph{symbolic differentiation}, \c{D} and \c{deriv}. They support arithmetic operations and functions on real numbers such as \c{sin}.
In \emph{MCMCglmm} (Monte Carl Markov Chain Generalised Linear Mixed Models), a \c{Dtensor} object is created, which contains expressions derived twice (with \c{DD}) and then typically in a list or a data frame.
\begin{lstlisting}
 expr<-expression(beta_1 + time*beta_2+u)
 mu = data.frame(beta_1=0.5, beta_2=1, time=3, u=2.3)
 D[i]<-eval(DD(expr, name[unlist(comb.pos[i,])]), mu)
\end{lstlisting}
% See https://github.com/cran/MCMCglmm/blob/cfac9e67ec73a3db2142826faae35e5d2318da31/man/evalDtensor.Rd

Package \c{tidyselect} is the backend of several \c{dplyr} and \c{tidyr} functions to select columns. It offers a DSL that makes it possible to select a set of columns, over a range of columns, a complement, an intersection, an union, with a regular expression and so on. \eval is used to traverse selectors stored as slots of a dataframe.

>>>>

  \item \emph{Assignments \c{(<-)}} appear in the cases where a developer wants
    to create or update a binding in a specific environment where either the
    name or the value come from some expression. Most cases originate from
    trivial code generation where the assign expression is assembled using
    \c{parse} or \c{substitute}, often in a loop. For example, the
    \c{PopED::get\_rse} uses it to assign a number of variables in the current
    scope: \lstinline|eval(parse(text=paste(default_args[[i]]), "<-", i))|.

@@

  \item \emph{Function calls \c{(F(), F(X), F(F(X)))}} represent two very
    frequent patterns: \emph{code generation} and \emph{call reflection}. There
    are numerous cases for code generation.
    %
    One occurring usage (present also in the Kaggle scripts) is to generate
    ad-hoc calls to libraries such as \c{ggplot2} when one needs to
    parameterize data or aesthetics selection from a variable. For example
    \lstinline|ggplot(df, aes(x=A, y=B))| maps X and Y axes in the plot to
    \c{A} and \c{B} columns in \c{df}. It might not be immediately
    clear\footnote{\cf
    \url{https://stackoverflow.com/questions/22309285/how-to-use-a-variable-to-specify-column-name-in-ggplot}}
    for a ggplot2 user how to modify the call to assign the axis columns from
    variables. Another case is the traditional use of \eval as a mean to reduce
    boilerplate code; simple and repetitive code can easily be replaced with
    judicious use of \eval. For example, the \c{data.table} package uses \eval
    to call the \c{options} function with named arguments taken from a vector
    of strings:

    \begin{lstlisting}[gobble=2]
    opts = c("datatable.verbose"="FALSE", ... )
    for (i in names(opts)) eval(parse(text=paste0("options(",i,"=",opts[i],")")))
    \end{lstlisting}

    Call reflection refers to the combination of \eval with \c{match.call}, a
    function that returns the current call with all its arguments captured as
    expressions. This is used mostly in the following circumstances: (1) to
    record a call for a later use in a possibly different environment, (2) pass
    most of the captured call arguments to another function or (3) evaluate
    arguments of the function in different environments. This is widely used in
    relation with statistical modeling in combination with a model fitting
    function or with \c{model.frame} and related functions. For example the
    \c{survival::coxph} function\footnote{Function fitting a Cox proportional hazards
    regression model.} contains:
    %
    \begin{lstlisting}[gobble=2]
    Call <- match.call()
    tform <- Call[c(1,indx)]  # only keep the arguments we wanted
    tform[[1L]] <- quote(stats::model.frame)  # change the function called
    mf <- eval(tform, parent.frame())
    \end{lstlisting}

@@

\subsection{Domain Specific Language}

The ability to capture current call as a language object (\c{match.call()}),
extract the unevaluated expression from function arguments (\c{substitute()})
and \eval with first class environment makes the R toolkit for building DSLs.
There are numerous widely spread DSLs used in R. For example, \c{data.table}
package defines a DSL for data frame manipulation using the subset operator
\c{[}. This allowing one to write compact queries such as \c{flights[carrier ==
  "AA", .N, by = origin]}\footnote{The \c{data.table} subset operation is
defined as \c{DT[i, j, k]}, where \c{j} expression is used to calculate on a
subset/reorder of rows from \c{i} grouped by \c{k}.} which will compute the
number of flights done by American Airlines for each origin airport.
Essentially, it uses \c{substitute} to capture the operand's expressions,
transforms then appropriately and finally uses \eval to execute the query.

Another example is the \c{tidyselect} package that defines a DSL for selecting
columns in a data frame powering number of data manipulation operation in the
\c{data.table} main competitor, the \c{dply} package. It uses eval to build an
interpreter for evaluating column selectors such as \c{select(df,
starts\_with("length") \& !(name:mass))} which returns all columns in \c{df}
that starts with the string \c{length} and is not between \c{name} and
\c{mass} columns.

% TODO Pierre
For instance, base R packages \emph{stats} provides two operators for \emph{symbolic differentiation}, \c{D} and \c{deriv}. They support arithmetic operations and functions on real numbers such as \c{sin}.
In \emph{MCMCglmm} (Monte Carl Markov Chain Generalised Linear Mixed Models), a \c{Dtensor} object is created, which contains expressions derived twice (with \c{DD}) and then typically in a list or a data frame.
\begin{lstlisting}
 expr<-expression(beta_1 + time*beta_2+u)
 mu = data.frame(beta_1=0.5, beta_2=1, time=3, u=2.3)
 D[i]<-eval(DD(expr, name[unlist(comb.pos[i,])]), mu)
\end{lstlisting}
% See https://github.com/cran/MCMCglmm/blob/cfac9e67ec73a3db2142826faae35e5d2318da31/man/evalDtensor.Rd
